Flatband ferromagnetism and dispersion relations of spin-1 excitations in topological Hubbard models

### 2018-06-13 paper/paper.tex
Version 1.0.

\begin{document}
\title{Flatband ferromagnetism and dispersion relations of spin-1 excitations in topological Hubbard models}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction strength upon which the ferromagnetic phase is stable, and elaborate the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, \textcolor{red}{and the scatterings between the Stoner continuum at these dips and the collective modes significantly reduce the latter's energy}, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.
\end{abstract}
\maketitle

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of an flat electron band with its filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state was also studied \cite{KA_PRL1994,SGDL_arXiv2008} and itinerant topological magnons has been reported quite recently \cite{SGDL_arXiv2008}.

@@

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical Hubbard interaction strength is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, upon which the ferromagnetic phase is stable. Furthermore the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, \textcolor{red}{and the scatterings between the Stoner continuum at these dips and the collective modes significantly reduce the latter's energy}, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.

@@

\subsection{Exact diagonalization with projection}\label{exact_diag_proj}
\par Exact diagonalization method with a projection onto the low-energy Hilbert space bas been widely applied to systems that host flat or nearly-flat energy bands \cite{NSCM_PRL2011,RB_PRX2011,NSRCM_PRB2011,SGDL_arXiv2008,NSRCM_PRL2012,DG_PRB2015}. This approach ignores the band mixing of electrons and applies when the Coulomb interaction is small compared to the energy gap of the electron bands.

@@

Here $d^{\dagger}_{\mathbf{k}_{i}-\mathbf{q}\downarrow}$ creates a spin-down electron with momentum $\mathbf{k}_{i}-\mathbf{q}$ and $d_{\mathbf{k}_{i}\uparrow}$ annihilates a spin-up electron with momentum $\mathbf{k}_{i}$. Therefore this basis labels a spin-1 scattering channel with the index $\mathbf{k}_{i}$. Thus the dimension of this Hilbert space scales linearly with respect to the number of electron momentums \cite{SGDL_arXiv2008}, and a much larger system can be numerically accessed than the usual exact diagonalization without projection, which enables us to analyze the properties of the spin-1 excitation spectra at the whole first Brillouin zone rather than some restricted discrete points solely. The matrix element of the projected Hamiltonian on this spin-1 excitation basis can be easily obtained after some algebra,

@@

$h^\text{flat}_{\mathbf{k}\sigma}$ shares the same eigenfunctions with $h_{\mathbf{k}\sigma}$ but has exactly-flat energy bands with the eigenvalues being $\pm 1$. $h^\text{flat}_{\mathbf{k}\sigma}$, together with the Hubbard interaction $H_U$, defines the flatband limit of the original Hamiltonian. To approach this limit, long-range hopping terms in the real space must be included \cite{NSCM_PRL2011}. To elaborate the physics related to the nonflatness of the $d$ band, we also calculated the spin-1 excitation spectra in the flatband limit for comparison. This can be done by simply ignoring the $\left[\varepsilon_d(\mathbf{k}_i-\mathbf{q})-\varepsilon_d(\mathbf{k}_i)\right]\delta_{\mathbf{k}_j,\mathbf{k}_i}$ term in Eq. (\ref{PHP}), because of the same eigenfunctions shared by $h^\text{flat}_{\mathbf{k}\sigma}$ and $h_{\mathbf{k}\sigma}$.

@@

\par The phase diagrams of the quarter-filled Chern Hubbard model and $Z_2$ Hubbard model are shown in Fig. \ref{phase}. Here, the colormap represents the ratio of the Hubbard interaction strength $U$ over the lower electron bandwidth $W$. FM denotes the ferromagnetic phase and NFM denotes the non-ferromagnetic phase. As is discussed in Sec. \ref{exact_diag_proj}, the phase boundary is determined as the onset point of $U$ below which negative spin-1 excitation spectra appear. It is clear that a critical Hubbard interaction strength is needed to maintain the ferromagnetically ordered ground state when the electron band has finite nonflatness. This observation is expected as a result from the competition between the kinetic energy and the potential energy of the electrons, as a fully spin-polarized state minimizes the energy of Hubbard interactions but cost more energy when the electron band disperses. An intriguing fact about the phase diagrams is that the phase boundaries always lie near the contour line with $U/W=2.0$.

\subsection{Chern Hubbard model}
\begin{figure}
\includegraphics[width=\columnwidth]{fmcispectrum}
\caption{(Color online) (a) Spin-1 excitation spectra of the Chern Hubbard model in the FM phase. Green solid lines denote the spin waves and the grey regions denote the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Blue solid lines denote the upper and lower boundaries of the Stoner continuum determined by the split band picture shown in (c). Black arrows mark the local minima of the Stoner continuum. (b) Spectral weights for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(0,0)$. (c) Illustration of the splitting of the lower electron bands for up spins (blue solid line) and down spins (blue dashed line) in the presence of ferromagnetism of the Chern Hubbard model. Inset shows the positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the dips of the Stoner continuum marked in (a). (d) Differences of the spectral weights subtracted by those in the flatband limit for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(\pi,0)$. The parameters are $t_1=1.0$, $t_2=0.49$ and $U=2.3$.}
\label{fmcispectrum}
\end{figure}

\par To gain a comprehensive understanding of the physics observed above, the spin-1 excitation spectra along a high symmetry path in the first Brillouin zone of the quarter-filled Chern Hubbard model with two different $t_2$ values are plotted in Fig. \ref{cispectrum}(a)-(b) as the green solid lines and the shaded areas. For a comparison, the spectra for those in the corresponding flat-band limit are shown as the green dashed lines as well. Apparently, for both the original model and the flat-band limit model, the spectra consist of two types of modes: the low-lying ones exhibiting well-defined band structures and the high-energy ones forming a continuum. The former are identified as the magnons with an acoustic branch and an optical branch while the latter are identified as the Stoner continuum. At $\Gamma$ point, the acoustic magnon band is gapless, which is the character of a ferromagnetic excitation with the Goldstone mode as the result of the spontaneous spin $SU(2)$ symmetry breaking. Remarkably,

@@

%\par The Hamiltonian has SU(2) spin rotation symmetry. The ferromagnetic ground state breaks SU(2) spin rotation symmetry to U(1) symmetry, and one would therefore expect the presence of a Goldstone mode in the spin-wave excitation spectrum due to spontaneous broken symmetry. We show such a mode at the center of the Brillouin Zone [see Fig.~\ref{fig3}]. The lowest two bands (black solid line and red dashed line) respectively corresponds to the acoustic and optical magnon (spin-wave excitation) bands degenerating along the $X-M$ path in the Brillouin Zone while the higher excitations represent to Stoner continuum in Fig.~\ref{fig3}. $G_{1}(\mathbf{k}_{i},\mathbf{Q})$ dominates the Stoner excitation gap while $G_{2}(\mathbf{k}_{i},\mathbf{k}_{j},\mathbf{Q})$ has very little contribution to the Stoner gap. With the increase of $W/U$, the optical spin-wave excitation is eventually absorbed in the Stoner continuum along the $\Gamma-X$ path in the Brillouin Zone [see Fig. \ref{fig3}(a),(b),(c)]. For $U_{c}=2t_{1}$ ($W/U_{c}=0.414$), there is a critical situation of ferromagnetic ground state, consistent with the theory of Tasaki \cite{Tasaki1} [see Fig.~\ref{fig3}(c)]. When $W/U>W/U_{c}$, the spin-wave excitation appears to be negative somewhere, and the ground state is no longer fully spin-polarized ferromagnetic state.  In order to investigate the effect of the $c$ band to the spin-1 excitation spectrum, we have to consider the case that the $c$ band is assumed as a completely flat-band. In this case, the excitation energies are proportional to the on-site repulsion $U$ instead of the nearest-neighbor hopping energy $t_{1}$ and the boundary of Stoner continuum coalesces into a line [see Fig.~\ref{fig3}(d)]. No matter how small the repulsion $U$ is, the ground state is a fully spin-polarized ferromagnetic state. In contrast to this case, the dispersion of the $c$ band makes the excitation spectrum have the downwarping behavior and gives rise to the phase transition for $U=2t_{1}$ [see Fig.~\ref{fig3}(a),(b),(c)]. The spin-wave excitation spectrum in Fig.~\ref{fig3}(d) is qualitatively similar to the result of the bosonization formalism in ref.~\cite{Doretto}. However, Our result takes into account the effect of the spin-wave interaction in the bosonization formalism. The bosonization formalism neglects the contribution of the nearly flat-band to the spin wave excitation and can't determine what the $W/U$ value is, the ground state is the fully spin-polarized ferromagnetic state.

\subsection{Topological Hubbard model}
\begin{figure}
\includegraphics[width=\columnwidth]{tispectrum}
\caption{(Color online) (a) and (b): Spin-1 excitation spectra of the $Z_2$ Hubbard model. Green solid lines denote the spin waves and the grey regions denote the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Black arrows mark the minima of the Stoner continuum. (c) and (d): Illustration of the splitting of the lower electron bands of the $Z_2$ Hubbard model. Insets show the positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the minima of the Stoner continuum marked in (a) and (b). $t_2=0.65$ for (a) and (c) and $t_2=0.741$ for (b) and (d). Other parameters are fixed at $t_1=1.0$, $U=2.3$.}
\label{tispectrum}
\end{figure}

%\par The spin-1 excitation spectrum of the the ferromagnetic phase of the topological Hubbard model is shown in Fig.~\ref{fig4}. This result indicates that the ground state of the topological Hubbard model is indeed the fully spin-polarized ferromagnetic state due to the fact that the flipping a spin from such state requires a positive energy for $U_{c}>1.82t_{1}$.

>>>>

\begin{document}
\title{Flatband ferromagnetism and dispersion relations of spin-1 excitations in topological Hubbard models}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction strength upon which the ferromagnetic phase is stable, and elaborate the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.
\end{abstract}
\maketitle

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of an flat electron band with its filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state was also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons has been reported quite recently \cite{SGDL_PRB2018}.

@@

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical Hubbard interaction strength is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, upon which the ferromagnetic phase is stable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.

@@

\subsection{Exact diagonalization with projection}\label{exact_diag_proj}
\par Exact diagonalization method with a projection onto the low-energy Hilbert space bas been widely applied to systems that host flat or nearly-flat energy bands \cite{NSCM_PRL2011,RB_PRX2011,NSRCM_PRB2011,SGDL_PRB2018,NSRCM_PRL2012,DG_PRB2015}. This approach ignores the band mixing of electrons and applies when the Coulomb interaction is small compared to the energy gap of the electron bands.

@@

Here $d^{\dagger}_{\mathbf{k}_{i}-\mathbf{q}\downarrow}$ creates a spin-down electron with momentum $\mathbf{k}_{i}-\mathbf{q}$ and $d_{\mathbf{k}_{i}\uparrow}$ annihilates a spin-up electron with momentum $\mathbf{k}_{i}$. Therefore this basis labels a spin-1 scattering channel with the index $\mathbf{k}_{i}$. Thus the dimension of this Hilbert space scales linearly with respect to the number of electron momentums \cite{SGDL_PRB2018}, and a much larger system can be numerically accessed than the usual exact diagonalization without projection, which enables us to analyze the properties of the spin-1 excitation spectra at the whole first Brillouin zone rather than some restricted discrete points solely. The matrix element of the projected Hamiltonian on this spin-1 excitation basis can be easily obtained after some algebra,

@@

$h^\text{flat}_{\mathbf{k}\sigma}$ shares the same eigenfunctions with $h_{\mathbf{k}\sigma}$ but has exactly-flat energy bands with the eigenvalues being $\pm 1$. $h^\text{flat}_{\mathbf{k}\sigma}$, together with the Hubbard interaction $H_U$, defines the flatband limit of the original Hamiltonian. To approach this limit, long-range hopping terms in the real space must be included \cite{NSCM_PRL2011}. To reveal the physics related to the nonflatness of the $d$ band, we also calculated the spin-1 excitation spectra in the flatband limit for comparison. This can be done by simply ignoring the $\left[\varepsilon_d(\mathbf{k}_i-\mathbf{q})-\varepsilon_d(\mathbf{k}_i)\right]\delta_{\mathbf{k}_j,\mathbf{k}_i}$ term in Eq. (\ref{PHP}), because of the same eigenfunctions shared by $h^\text{flat}_{\mathbf{k}\sigma}$ and $h_{\mathbf{k}\sigma}$.

@@

\par Before the detailed discussion on the spin-1 excitation spectra of the quarter-filled Chern Hubbard model and $Z_2$ Hubbard model, we present their phase diagrams first, which are shown in Fig. \ref{phase}. Here, the colormap represents the ratio of the Hubbard interaction strength $U$ over the lower electron bandwidth $W$. FM denotes the ferromagnetic phase and NFM denotes the non-ferromagnetic phase. As is discussed in Sec. \ref{exact_diag_proj}, the phase boundary is determined as the onset point of $U$ below which negative spin-1 excitation spectra arise. The phase diagrams of these two models appear quite similar. It is obvious that a critical Hubbard interaction strength is needed to maintain the ferromagnetically ordered ground state when the electron band has finite nonflatness. Furthermore the phase boundaries always lie near the contour line with $U/W=2.0$. Qualitatively speaking, these observations are expected as a result from the competition between the kinetic energy and the potential energy of the electrons, as a fully spin-polarized state minimizes the energy of Hubbard interactions but cost more energy when the electron band disperses. In the following, we will elaborate the spin-1 excitation spectra of these models to investigate the physics observed above microscopically and quantitatively.

\subsection{Chern Hubbard model}\label{CHM}
\begin{figure}
\includegraphics[width=\columnwidth]{fmcispectrum}
\caption{(Color online) (a) Spin-1 excitation spectra of the Chern Hubbard model in the FM phase. Green solid lines denote the spin waves and the grey region denotes the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Blue solid lines denote the upper and lower boundaries of the Stoner continuum determined by the split band picture shown in (c). Black arrows mark the local minima of the Stoner continuum. (b) Spectral weights for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(0,0)$. (c) Illustration of the splitting of the lower electron bands for up spins (blue solid line) and down spins (blue dashed line) in the presence of ferromagnetism of the Chern Hubbard model. Inset shows the positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the dips of the Stoner continuum marked in (a). (d) Differences of the spectral weights subtracted by those in the flatband limit for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(\pi,0)$. The parameters are $t_1=1.0$, $t_2=0.49$ and $U=2.3$.}
\label{fmcispectrum}
\end{figure}

\par In this subsection, we focus on the quarter-filled Chern Hubbard model. Its spin-1 excitation spectra in the FM phase along a high symmetry path in the first Brillouin zone are plotted in Fig. \ref{fmcispectrum}(a) as two low-lying green solid lines and a high-energy shaded area, with the parameters marked by the red star in Fig. \ref{phase}(a). The modes in the green solid lines and the shaded area take on distinct behaviors: the former exhibits well-defined band structures while the latter forms a continuum. In fact, for these two kinds of modes, the patterns of the contributions from each spin-1 particle-hole scattering channel of electrons, which are embodied in the eigenvectors $\Psi_{\mathbf{q}}(\mathbf{k}_i)$ of Eq. \ref{PHP}, are quite different. In Fig. \ref{fmcispectrum}(b), the spectral weights of each scattering channel, i.e. $|\Psi_{\mathbf{q}}(\mathbf{k}_i)|^2$ as a function of $\mathbf{k}_i$, for the four lowest eigen levels with the center-of-mass momentum $\mathbf{q}=(0,0)$ ($\Gamma$ point) are shown. It is clear that for the modes in the green solid lines [Fig. \ref{fmcispectrum}($\text{b}_1$) and Fig. \ref{fmcispectrum}($\text{b}_2$)], the spectral weights come from a quite broad range of scattering channels; indeed, the spectral weight for the lowest eigen mode at $\Gamma$ point is even homogeneous for all scattering channels. However, for those in the shaded area [Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4$)], the spectral weights almost come from a single scattering channel. Therefore, the former modes are collective spin-1 excitations and are identified as the spin waves while the latter modes are individual spin-1 excitations and are identified as the Stoner continuum. Due to the $AB$ sublattice of the model, the spin waves contains an acoustic branch and an optical branch. The acoustic branch is gapless, which is consistent with the Goldstone theorem as the spin-fully polarized ground state spontaneously breaks the spin $SU(2)$ rotation symmetry.

\par To explore the physics related to the nonflatness of the lower electron band, the corresponding spin-1 excitation spectra in the flatband limit are also shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines. The flatband-limit spectra consist of two parts as well, i.e. the low-lying spin waves and the high-energy Stoner continuum. However, in this case, the Stoner continuum is just a pile of featureless flat lines at $U/2$. Comparing the spectra of the original model with this, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a), we can conclude that the nonflatness of the electron band introduces dips of the lower boundary of the Stoner continuum. Furthermore, around the valleys of these dips, the energies of the collective modes are apparently renormalized downward, leading to the occurrence of roton-like local minima of the optical branch.

\par The above physics can be illustrated by a simple band-splitting picture as shown in Fig. \ref{fmcispectrum}(c). Roughly speaking, in the presence of ferromagnetism, the spin-down electron band and spin-up electron band are split by an energy gap proportional to $U\langle m \rangle$, where $U$ is the Hubbard interaction strength and $\langle m \rangle$ is the average magnetic moment per site. For a quarter-filled electron model with two inequivalent sublattices, $\langle m \rangle=1/2$. Thus the energy splitting is $U/2$. Then an individual spin-1 excitation, or a mode in the Stoner continuum, can be viewed as the excitation of an electron from the lower fully-filled spin-up band to the upper empty spin-down band. This argument can be verified by our numerical results. In fact, the sharp locality on the $\mathbf{k}_i$ space of the spectral weights of a mode in the Stoner continuum [as has be seen in Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4)$] implies that this kind of mode is mainly determined by the corresponding diagonal term in Eq. \ref{PHP}. By numerics, we also find that the term $\frac{U}{N} \sum_{a=1,2} \sum_{\mathbf{p}\neq\mathbf{k}_i} \left| \mu_{a,\mathbf{p}\uparrow} \right|^2 \left| \mu_{a,\mathbf{k}_{i}-\mathbf{q}\downarrow} \right|^2 \sim \frac{U}{2}$. Therefore, the excitation energies of the Stoner continuum can be approximated by $\varepsilon_{d\downarrow} (\mathbf{k}_i-\mathbf{q}) + \frac{U}{2} -\varepsilon_{d\uparrow} (\mathbf{k}_i)$, which is consistent with the above picture. In Fig. \ref{fmcispectrum}(a), the lower and upper boundaries of the Stoner continuum approximated by this argument are plotted as the blue solid lines, which fits extremely well with the true boundaries of the shaded area, thus verifies the validity of this simple argument. Now we can find out the dominant scattering channels that contribute most to the dips of the Stoner continuum and read off their center-of-mass momentums easily. In the inset of Fig. \ref{fmcispectrum}(c), we mark the local maxima of the spin-up band with red solid circles and the local minima of the spin-down band with purple ones. The scattering channels corresponding to the valley bottoms marked by the black arrows in Fig. \ref{fmcispectrum}(a) are also indicated by green arrows. Obviously, a spin-1 excitation of an electron from one of the spin-up maxima to one of the spin-down minima appears as a valley bottom of the lower boundary of the Stoner continuum and the momentum difference of these two electron state provides the center-of-mass momentum of that spin-1 excitation.

\par Within this picture, the collective modes of the spectra can be viewed as bound states in the gap of the Stoner continuum. Then the physics of the downward renormalization of the energies of the collective modes induced by the dips of the Stoner continuum is quite similar to that observed in the two-dimensional antiferromagents, where the suppression of the energies of the spin waves are explained by their couplings to the nearby deconfined spinon continuum \cite{ZFSMC_PRL2006,TS_PRL2013,DMCNTPEMIR_NP2015,YWDYL_arXiv2018}. To understand the renormalization of the collective modes observed here, in Fig. \ref{fmcispectrum}(d), we plot the spectral weights of the four lowest eigen levels subtracted by those in the flatband limit with the center-of-mass momentum $q=(\pi,0)$ ($X_1$ point). The differences concentrate around the scattering channels labeled by $\mathbf{k}_i=(\pm\pi,0)$ and $\mathbf{k}_i=(0,\pm\pi)$. However, the results for the collective modes [Fig. \ref{fmcispectrum}($\text{d}_1$) and Fig. \ref{fmcispectrum}($\text{d}_2$)] are positive while those for the individual modes at the valley bottom of the Stoner continuum [Fig. \ref{fmcispectrum}($\text{d}_3$) and Fig. \ref{fmcispectrum}($\text{d}_4$)] are negative. Therefore a noticeable amount of spectral weights transfer from the latter to the former, which indicates a strong coupling between the spin waves and the nearby Stoner continuum. This coupling renormalizes the energies of the spin waves and leads to the occurrence of a roton-like local minimum for the optical branch.

@@

\par The physics discussed above is essential for the destabilization of the ferromagnetic phase. In Fig. \ref{pbcispectrum}(a), we plot the spin-1 excitation spectra of the model at the FM/NFM phase boundary, with the parameters marked by the blue star in Fig. \ref{phase}(a). The corresponding illustration of the split-band picture is shown in Fig. \ref{pbcispectrum}(b). It can be seen that with this increased nonflatness, the valley bottom marked by the black arrow in Fig. \ref{pbcispectrum}(a) of the Stoner continuum goes deeper toward zero energy. Besides the roton-like local minimum on the optical band, a roton-like local minimum on the acoustic band also appears, which can be seen more clearly in the inset of Fig. \ref{pbcispectrum}(a). With the further increase of the nonflatness of the electron band, the energy of this newly formed roton mode will go down through the zero energy, which leads to the final destabilization of the ground state. From Fig. \ref{fmcispectrum}(c) and Fig. \ref{pbcispectrum}(b), we can see that the ferromagnetic ground state is quite robust against the spin-1 flips until the top of the spin-up band is approaching near to the bottom of the spin-down band. Considering that the energy splitting of these two bands is $U/2$, the destabilization point would be quite close to $U/2\sim W$.

\subsection{$Z_2$ Hubbard model}\label{ZHM}
\begin{figure}
\includegraphics[width=\columnwidth]{tispectrum}
\caption{(Color online) (a) and (b): Spin-1 excitation spectra of the $Z_2$ Hubbard model. Green solid lines denote the spin waves and the grey regions denote the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Black arrows mark the minima of the Stoner continuum. (c) and (d): Illustration of the splitting of the lower electron bands of the $Z_2$ Hubbard model. Insets show the positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the minima of the Stoner continuum marked in (a) and (b). $t_2=0.65$ for (a) and (c) and $t_2=0.741$ for (b) and (d). Other parameters are fixed at $t_1=1.0$, $U=2.3$.}
\label{tispectrum}
\end{figure}

\par In this subsection, we focus on the quarter-filled $Z_2$ Hubbard model. Its spin-1 excitation spectra along a high symmetry path in the first Brillouin zone with the parameters marked by the red star and blue star in Fig. \ref{phase}(b) are plotted in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b), respectively. As is similar to Sec. \ref{CHM}, the low-lying green solid lines represent the spin waves (collective modes) and the high-energy shaded areas represent the Stoner continuum (individual modes). A distinction from the Chern Hubbard model is that the collective modes here are gapped because the $Z_2$ Hubbard model explicitly breaks the spin $SU(2)$ rotation symmetry and no spontaneous continuous symmetry breaking occurs in the ferromagnetic phase. The corresponding spectra in the flatband limit are shown as the green dashed lines in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b) as well for comparison. Similarly, the nonflatness of the electron band introduces dips of the Stoner continuum, and the energies of the spin waves around the valley bottoms of the Stoner continuum are renormalized downward, which results in roton-like modes on both branches of the spin waves. The dominant scattering channels leading to the dips of the Stoner continuum and the renormalization of the spin waves can also be determined by the previously proposed band-splitting picture, which is illustrated in Fig. \ref{tispectrum}(c) and Fig. \ref{tispectrum}(d). Their insets shows the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Here only the scattering channels that contribute to the valley bottoms marked by the black arrows in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b) are stressed because their induced roton-like modes lead to the destabilization of the ferromagnetic ground state [See Fig. \ref{tispectrum}(b)]. We want to remark that although the center-of-mass momentums of the forementioned valley bottoms are the same, the scattering channels are different because the bottom of the spin-down band changes from the $\Gamma$ and $M_{1,2,3,4}$ points to the $X_{1,2}$ and $Y_{1,2}$ points when $t_2$ changes from $0.65$ to $0.741$ [See Fig. \ref{tispectrum}(c) and Fig. \ref{tispectrum}(d)]. Last but not least, the ferromagnetic ground state is also quite robust against the spin flips until the bottom of the spin-down band almost overlaps with the spin-up band on energy, which means the phase boundary of the FM/NFM is also quite close to $U/W=2$.

\section{Summary and discussion}\label{sd}
\par In summary, we have considered the flatband ferromagnetic phases in a quarter-filled Chern Hubbard model and a quarter-filled $Z_2$ Hubbard model in this paper. By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction strength upon which the ferromagnetic phase is stable, and elaborate the ferromagnetic spin-1 excitation spectra of these models. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.

\par We want to make a comparison of our flatband-limit results with those obtained by the generalized bosonization formalism proposed in Ref. \cite{DG_PRB2015}, where both the nonflatness of the electron bands and the self-interactions between the magnons are ignored. For the Chern Hubbard model, our results are qualitatively consistent with theirs [See Fig. \ref{fmcispectrum}(a) in our paper and Fig. 4 in Ref. \cite{DG_PRB2015}]. However, for the $Z_2$ Hubbard model, our results are different from theirs in that the band bottom of the spin waves in our results lies at the $M_{1,2,3,4}$ points while theirs lies at the $\Gamma$ point [See Figs. \ref{tispectrum}(a)-(b) in our paper and Fig. 5 in Ref. \cite{DG_PRB2015}]. We attribute this discrepancy to the self-interactions of the magnons because our numerical method treat the system after the projection exactly. Although correlated nearly-flat topological bands are hard to realize in real materials, they are possible to be designed in cold-atom systems. In fact great progress has been made very recently \cite{LCJPS_N2009,AALBPB_PRL2013,MSKBK_PRL2013,JMDLUGE_N2014}. We thus expect our results to be tested in future cold-atom experiments.
